---
title: "Customer Segmentation-Phase2"
author: "Silvester Cheptumo"
date: "2023-07-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(RMySQL)
library(DBI)
library(dplyr)
library(lubridate)
library(factoextra)
library(data.table)
library(ggplot2)
library(caret)
library(tidyr)
library(tidyverse)
library(parallel)
```
```{r}
con<-dbConnect(MySQL(),
                user="Kaimugul",
                password="Kabon@123#",
                dbname="Banking",
                host="127.0.0.1")
```
Segmentation1
```{r}
#SEGMENT 1
credit_card_data<- dbGetQuery(con, paste0("SELECT cust_id, acct_id, card_no, 
                                                  transaction_date,transaction_amount, transaction_category
                          FROM credit_card_transactions"))
# Assuming 'transaction_date' is a character vector in the format "YYYY-MM-DD HH:MM:SS"
credit_card_data$transaction_date <- as.POSIXct(credit_card_data$transaction_date, format = "%Y-%m-%d %H:%M:%S")

credit_card_data$month <- lubridate::month(credit_card_data$transaction_date)
credit_card_data$year <- lubridate::year(credit_card_data$transaction_date)

# Step 2.1: Group on cardno, month, and category and get total month level purchase
monthly_purchase <- credit_card_data %>%
  group_by(card_no, transaction_category, month, year) %>%
  summarize(total_monthly_purchase = sum(transaction_amount))

# Step 2.2: Group on cardno and category and get the average of month level purchase from data in step 2.1
avg_monthly_purchase <- monthly_purchase %>%
  group_by(card_no, transaction_category) %>%
  summarize(avg_purchase = mean(total_monthly_purchase))

# Step 2.3: Convert step 2.2 into a feature matrix (using dcast)

feature_matrix <- dcast(avg_monthly_purchase, card_no ~ transaction_category, value.var = "avg_purchase")

# Replace NA values with zero
feature_matrix[is.na(feature_matrix)] = 0

#normalizing and standardizing the data 
min_max_normalized_data <- as.data.frame(apply(feature_matrix, 2, function(x) (x - min(x)) / (max(x) - min(x))))

#using the elbow method to find the optimal number of clusters
wcss<-numeric(10)
for (k in 1:10){
  kmeans_model<-kmeans(min_max_normalized_data, centers=k)
  wcss[k]<-kmeans_model$tot.withinss
}
plot(1:10, wcss, type = "b", pch=19, frame=FALSE, xlab = "Number of Clusters(k)", 
     ylab = "Within-cluster Sum of Squares(WCSS)")

# Step 2.4: 40 clusters on the feature matrix (exclude cardno when building it)
library(cluster)
set.seed(123)
num_clusters <- 4
kmeans_model <- kmeans(min_max_normalized_data, centers = num_clusters)


# Create a new column "cluster" and assign cluster labels
feature_matrix$num_cluster<-kmeans_model$cluster

#aggregatting spending by clusters
# Calculate the total spending in each cluster
cluster_summary <- feature_matrix %>%
  group_by(num_cluster) %>%
  summarize(
    no_of_card_customers = n(),
    total_spending = sum(accomodation, bill_payments, cosmetics, electronics,
                         entertainment, fashion, food_kiosk, fuel)
  )
# Step 2.6: For each cluster, calculate the number of card customers, and average of other columns
cluster_summary <- feature_matrix %>%
  group_by(num_cluster) %>%
  summarize(
    no_of_card_customers = n(),
    avg_purchase_category_accomodation = mean(accomodation),
    avg_purchase_category_bill_payments = mean(bill_payments),
    avg_purchase_category_cosmetics = mean(cosmetics),
    avg_purchase_category_electronics = mean(electronics),
    avg_purchase_category_entertaiment = mean(entertaiment),
    avg_purchase_category_fashion = mean(fashion),
    avg_purchase_category_food_kiosk = mean(food_kiosk),
    avg_purchase_category_fuel = mean(fuel)
  )
#aggregating the clusters 
cluster_summary_aggregate <- cluster_summary %>%
  mutate(
    total_spending = avg_purchase_category_accomodation +
                     avg_purchase_category_bill_payments +
                     avg_purchase_category_cosmetics +
                     avg_purchase_category_electronics +
                     avg_purchase_category_entertaiment +
                     avg_purchase_category_fashion +
                     avg_purchase_category_food_kiosk +
                     avg_purchase_category_fuel
  )
#naming my clusters 
# Name the clusters based on their total spending
cluster_summary <- cluster_summary %>%
  mutate(
    cluster_name = case_when(
      num_cluster == 3 ~ "Highest Spenders",
      num_cluster == 4 ~ "High Spenders",
      num_cluster == 1 ~ "Medium spenders",
      num_cluster == 2 ~ "Low Spenders"
    )
  )
cluster_summary
#segmentation of the clusters
```
```{r}
str(cluster_summary)
```
Visualizing clusters for segmentation1
#working with 4 clusters
#visualize the clusters

```{r}
# Reshape the data into long format for plotting
cluster_summary_long <- cluster_summary %>%
  pivot_longer(cols = starts_with("avg_purchase_category_"),
               names_to = "purchase_category",
               values_to = "avg_percentage")

# Define custom cluster names
custom_cluster_names <- c("low Spend", "Medium-Low Spend", "High Spend", "Medium-High Spend")

# Visualize the average purchase percentages for each cluster and purchase category
ggplot(cluster_summary_long, aes(x = purchase_category, y = avg_percentage, fill = as.factor(num_cluster))) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_discrete(name = "Cluster",
                      breaks = unique(cluster_summary_long$num_cluster),
                      labels = custom_cluster_names) +
  labs(title = "Average Purchase Percentages by Cluster and Purchase Category",
       x = "Purchase Category",
       y = "Average Percentage") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



Segmentation 2: Build customer segment based on purchases, categories & payments based on debit card transactions

```{r}
#step1
debit_card_data<- dbGetQuery(con, paste0("SELECT cust_id, acct_id, card_no, 
                                                  transaction_date,transaction_amount, transaction_category
                          FROM debit_card_transactions"))
debit_card_data
# Assuming 'transaction_date' is a character vector in the format "YYYY-MM-DD HH:MM:SS"
debit_card_data$transaction_date <- as.POSIXct(debit_card_data$transaction_date, format = "%Y-%m-%d %H:%M:%S")

debit_card_data$month <- lubridate::month(debit_card_data$transaction_date)
debit_card_data$year <- lubridate::year(debit_card_data$transaction_date)

#Step2
# Step 2.1: Group on cardno, month, and category and get total month level purchase
monthly_purchase <- debit_card_data %>%
  group_by(card_no, transaction_category, month, year) %>%
  summarize(total_monthly_purchase = sum(transaction_amount))

# Step 2.2: Group on cardno and category and get the average of month level purchase from data in step 2.1
avg_monthly_purchase <- monthly_purchase %>%
  group_by(card_no, transaction_category) %>%
  summarize(avg_purchase = mean(total_monthly_purchase))

# Step 2.3: Convert step 2.2 into a feature matrix (using dcast)

feature_matrix <- dcast(avg_monthly_purchase, card_no ~ transaction_category, value.var = "avg_purchase")

# Replace NA values with zero
feature_matrix[is.na(feature_matrix)] = 0

#normalizing and standardizing the data 
min_max_normalized_data <- as.data.frame(apply(feature_matrix, 2, function(x) (x - min(x)) / (max(x) - min(x))))

#using the elbow method to find the optimal number of clusters
wcss<-numeric(10)
for (k in 1:10){
  kmeans_model<-kmeans(min_max_normalized_data, centers=k)
  wcss[k]<-kmeans_model$tot.withinss
}
plot(1:10, wcss, type = "b", pch=19, frame=FALSE, xlab = "Number of Clusters(k)", 
     ylab = "Within-cluster Sum of Squares(WCSS)")

# Step 2.4: 40 clusters on the feature matrix (exclude cardno when building it)
library(cluster)
set.seed(123)
num_clusters <- 4
kmeans_model <- kmeans(min_max_normalized_data, centers = num_clusters)


# Step 2.5: Attach cluster number back to min_max_normalized_data
feature_matrix$num_clusters <- kmeans_model$cluster


# Step 2.6: For each cluster, calculate the number of card customers, and average of other columns
cluster_summary2 <- feature_matrix %>%
  group_by(num_clusters) %>%
  summarize(
    no_of_card_customers = n(),
    avg_purchase_category_accomodation = mean(accomodation),
    avg_purchase_category_bill_payments = mean(bill_payments),
    avg_purchase_category_cosmetics = mean(cosmetics),
    avg_purchase_category_electronics = mean(electronics),
    avg_purchase_category_entertaiment = mean(entertaiment),
    avg_purchase_category_fashion = mean(fashion),
    avg_purchase_category_food_kiosk = mean(food_kiosk),
    avg_purchase_category_fuel = mean(fuel),
    avg_purchase_category_restaurent=mean(restaurent),
    avg_purchase_category_salon_spa=mean(salon_spa),
    avg_purchase_category_retail_outlet=mean(retail_outlet)
  )
cluster_summary2

#Calculate the number of card customers, and average of other columns
cluster_summary2 <- feature_matrix %>%
  group_by(num_clusters) %>%
  summarize(
    no_of_card_customers = n(),
    avg_purchase_category_accomodation = mean(accomodation),
    avg_purchase_category_bill_payments = mean(bill_payments),
    avg_purchase_category_cosmetics = mean(cosmetics),
    avg_purchase_category_electronics = mean(electronics),
    avg_purchase_category_entertaiment = mean(entertaiment),
    avg_purchase_category_fashion = mean(fashion),
    avg_purchase_category_food_kiosk = mean(food_kiosk),
    avg_purchase_category_fuel = mean(fuel),
    avg_purchase_category_restaurent=mean(restaurent),
    avg_purchase_category_salon_spa=mean(salon_spa),
    avg_purchase_category_retail_outlet=mean(retail_outlet)
  )
cluster_summary2
#aggregating the clusters 
cluster_summary2 <- cluster_summary2 %>%
  mutate(
    total_spending = avg_purchase_category_accomodation +
      avg_purchase_category_bill_payments +
      avg_purchase_category_cosmetics +
      avg_purchase_category_electronics +
      avg_purchase_category_entertaiment +
      avg_purchase_category_fashion +
      avg_purchase_category_food_kiosk +
      avg_purchase_category_fuel +
      avg_purchase_category_retail_outlet +
      avg_purchase_category_salon_spa +
      avg_purchase_category_restaurent
  )
cluster_summary2

#naming the 4 clusters according to their total spending
cluster_summary2 <- cluster_summary2 %>%
  mutate(
    cluster_name = case_when(
      num_clusters == 4 ~ "Top Spenders",
      num_clusters == 1 ~ "High Spenders",
      num_clusters == 3 ~ "Medium spenders",
      num_clusters == 2 ~ "Low spenders"
    ))
cluster_summary2
#plot for the data points in the segment 

  library(plotly)
  pie_chart <- plot_ly(cluster_summary2, labels = ~paste(cluster_name),
                       values = ~no_of_card_customers, type = "pie",
                       textinfo = "label+percent",
                       textposition = "inside", insidetextfont = list(color = '#FFFFFF'))
  
  pie_chart %>% layout(title = "Number of Customers per Cluster")


```

```


Segmentation3:Build customer segment based on transactions through internet banking

```{r}
core_data<- dbGetQuery(con, paste0("SELECT cust_id,
                                                  trans_date, amount, service
                          FROM core_transactions"))

internet_banking_data<-subset(core_data, service=="internet_banking")
internet_banking_data

# Assuming 'transaction_date' is a character vector in the format "YYYY-MM-DD"
internet_banking_data$trans_date <- as.POSIXct(internet_banking_data$trans_date, format = "%Y-%m-%d")

internet_banking_data$month <- lubridate::month(internet_banking_data$trans_date)
internet_banking_data$year <- lubridate::year(internet_banking_data$trans_date)

internet_banking_data

# Calculate average transaction amount and total number of transactions for each customer
customer_summary <- internet_banking_data%>%
  group_by(cust_id,year, service) %>%
  summarize(avg_internet_transactions = mean(amount),
    total_transactions = n())
customer_summary

# Perform min-max scaling for the features
numeric_cols<-customer_summary[,c("avg_internet_transactions", "total_transactions")]


numeric_cols_normalized <- as.data.frame(apply(numeric_cols, 2, function(x) (x - min(x)) / (max(x) - min(x))))


# elbow method to find optimal clusters
wcss <- numeric(10)
for (k in 1:10) {
  kmeans_model <- kmeans(numeric_cols_normalized, centers = k)
  wcss[k] <- kmeans_model$tot.withinss
}

plot(1:10, wcss, type = "b", pch = 19, frame = FALSE, xlab = "Number of Clusters (k)", 
     ylab = "Within-cluster Sum of Squares (WCSS)")


# running the kmeans algorithmn 
library(cluster)
set.seed(123)
No_clusters <- 4
kmeans_model3 <- kmeans(numeric_cols_normalized, centers = No_clusters)
kmeans_model3


# Step 2.5: Attach cluster number back to min_max_normalized_data
customer_summary$No_clusters <- kmeans_model3$cluster
str(customer_summary)

# Create a data frame with the cluster summaries
cluster_summary3 <- customer_summary %>%
  group_by(No_clusters) %>%
  summarize(
    no_of_card_customers = n(),
    avg_internet_transactions = mean(avg_internet_transactions),
    count_transactions = sum(total_transactions))
str(cluster_summary3)

#naming the clusters based on the number of transactions per cluster
cluster_summary3_naming <- cluster_summary3 %>%
  mutate(
    cluster_name = case_when(
      No_clusters == 1 ~"Moderate internet banking users",
      No_clusters == 2 ~"High internet banking users",
      No_clusters == 3 ~"Highest internet banking users",
      No_clusters == 4 ~"Lowest internet banking users"
    )
  )
cluster_summary3_naming

#plotting a piechart showing number of customers per cluster
library(plotly)
pie_chart3 <- plot_ly(cluster_summary3_naming, labels = ~paste(cluster_name),
                       values = ~count_transactions, type = "pie",
                       textinfo = "label+percent",
                       textposition = "inside", insidetextfont = list(color = '#FFFFFF'))
  
pie_chart3 %>% layout(title = "Number of Customers per Cluster")
```




Segmentation4: Build customer segment based on demographics

```{r}
customer_demographics<- dbGetQuery(con, paste0("SELECT cust_id, gender, date_of_birth, occupation,state, 
                                   education, monthly_income
                                   FROM savings_accounts"))
#calculating the age of the customers
customer_demographics$Age <- year(today()) - year(customer_demographics$date_of_birth)
customer_demographics$date_of_birth <- customer_demographics$age

# Encoding my data
encoded_data <- model.matrix(~gender + education + occupation -1, customer_demographics)
encoded_data

#scaling the numerical data
numeric_cols1 <- customer_demographics[, c("Age", "monthly_income")]

min_max_normalized4 <- as.data.frame(apply(numeric_cols1, 2, function(x) (x - min(x)) / (max(x) - min(x))))

#binding the columns
binded_customer_demographics <- bind_cols(min_max_normalized4, encoded_data)
binded_customer_demographics

#clustering using kmeans
wcss <- numeric(10)
for (k in 1:10) {
  kmeans_model <- kmeans(binded_customer_demographics, centers = k)
  wcss[k] <- kmeans_model$tot.withinss
}

plot(1:10, wcss, type = "b", pch = 19, frame = FALSE, xlab = "Number of Clusters (k)", 
     ylab = "Within-cluster Sum of Squares (WCSS)")

#kmeans clustering
library(cluster)
set.seed(123)
num_clusters <- 4
kmeans_model4 <- kmeans(min_max_normalized4, centers = num_clusters)

#attaching the clusters to the dataset
customer_demographics$num_clusters <- kmeans_model4$cluster
customer_demographics

# Create a data frame with the cluster summaries
cluster_summary4 <- customer_demographics %>%
  group_by(num_clusters)
str(cluster_summary4)

# Calculate the number of customers per cluster
customers_per_cluster <- cluster_summary4 %>%
  count(num_clusters)

# Join customers_per_cluster with cluster_summary4
cluster_summary4 <- left_join(cluster_summary4, customers_per_cluster, by = "num_clusters")


#Calculate the total monthly income per cluster
total_income_per_cluster <- cluster_summary4 %>%
  group_by(num_clusters) %>%
  summarise(total_income = sum(monthly_income, na.rm = TRUE))

# Calculate the average monthly income per cluster
average_income_per_cluster <- cluster_summary4 %>%
  group_by(num_clusters) %>%
  summarise(average_income = mean(monthly_income, na.rm = TRUE))

#naming the clusters based on the average monthly income per cluster
cluster_summary_naming <- cluster_summary4 %>%
  mutate(
    cluster_name = case_when(
      num_clusters == 1 ~"Medium Earners",
      num_clusters == 2 ~"High Earners",
      num_clusters == 3 ~"Top Earners",
      num_clusters == 4 ~"Least Earners"
    )
  )

```
```{r}
library(plotly)
  pie_chart <- plot_ly(cluster_summary_naming, labels = ~paste(cluster_name),
                       values = ~n, type = "pie",
                       textinfo = "label+percent",
                       textposition = "inside", insidetextfont = list(color = '#FFFFFF'))
  
  pie_chart %>% layout(title = "Number of Customers per Cluster")
```




Segmentation5: Build customer segment based on EOD (End of the day balances)

```{r}
EOD_balance<- dbGetQuery(con, paste0("SELECT cust_id, trans_date, amount, balance
                                   FROM eod_balance"))

EOD_balance$trans_date <- as.POSIXct(EOD_balance$trans_date, format = "%Y-%m-%d")

EOD_balance$month <- lubridate::month(EOD_balance$trans_date)
EOD_balance$year <- lubridate::year(EOD_balance$trans_date)

# isolating the columns to be normalized and doing the normalization
numeric_cols4 <- EOD_balance[,c("balance", "amount")]
normalized4 <- as.data.frame(apply(numeric_cols4, 2, function(x) (x - min(x)) / (max(x) - min(x))))

#visual plots for the clusters 
#clustering using kmeans
wcss <- numeric(10)
for (k in 1:10) {
  kmeans_model <- kmeans(normalized4, centers = k)
  wcss[k] <- kmeans_model$tot.withinss
}

plot(1:10, wcss, type = "b", pch = 19, frame = FALSE, xlab = "Number of Clusters (k)", 
     ylab = "Within-cluster Sum of Squares (WCSS)")
#from the above plot, the optimal clusters is 3

# Import the necessary libraries
library(cluster)

clusters <- 3
# Run the k-means clustering algorithm
kmeans_model <- kmeans(normalized4, centers = clusters)

# Get the cluster labels
cluster_labels <- kmeans_model$cluster

# Combine the cluster labels with the dataset
EOD_balance <- cbind(EOD_balance, cluster_labels)
str(EOD_balance)
#performing cluster summary
# Group the data by the cluster labels
EOD_balance <- EOD_balance %>% group_by(cluster_labels)

# Summarize the data within each cluster
EOD_balance_summary <- EOD_balance %>% summarize(
  count = n(),
  mean_balance = mean(balance)
)
str(EOD_balance_summary)
#naming the clusters according to the mean balance
EOD_balance_naming <- EOD_balance_summary %>%
  mutate(
    cluster_name = case_when(
      cluster_labels == 1 ~"Low Balance",
      cluster_labels == 2 ~"High Balance",
      cluster_labels == 3 ~"Medium Balance"
    )
  )
  str(EOD_balance_naming)
#piechart for the datapoints in the cluster summary
library(plotly)
# Create a pie chart
pie_chart <- plot_ly(
  data = EOD_balance_naming,
  labels = ~cluster_name,
  values = ~count,
  type = "pie",
  textinfo = "label+percent"
)

# Add a title to the pie chart
pie_chart <- pie_chart %>%
  layout(title = "Cluster Distribution")

# Display the pie chart
pie_chart

#bar graph of customer balance per cluster 
# Create a bar graph with rotated x-axis labels
# Create a bar graph with hover text labels
bar_graph <- ggplot(EOD_balance_naming, aes(x = cluster_name, y = mean_balance, label = scales::comma(mean_balance))) +
  geom_bar(stat = "identity", fill = "green") +
  geom_text(aes(label = scales::comma(mean_balance)), vjust = -0.5) +  # Add hover text labels
  labs(
    x = "Cluster Name",
    y = "Mean Customer Balance",
    title = "Customer Balance per Cluster"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels

# Display the bar graph
bar_graph


```



