---
title: "QueryingDatabase"
author: "Silvester Cheptumo"
date: "2023-07-16"
output: html_document


---
```{r}
library(DBI)
library(RMySQL)
```
```{r}
library(dplyr)
library(lubridate)

```
#Rule1SQL
```{r}
con<-dbConnect(MySQL(),
                user="Kaimugul",
                password="Kabon@123#",
                dbname="Banking",
                host="127.0.0.1")
# Query the MySQL table
query <- "SELECT *
          FROM core_transactions
          WHERE trans_date = '2021-10-21'
          AND amount > 1000000;"
result <- dbGetQuery(con, query)
print(result)

dbDisconnect(con)
```
#Rule1R language
```{R}
con<-dbConnect(MySQL(),
                user="Kaimugul",
                password="Kabon@123#",
                dbname="Banking",
                host="127.0.0.1")
#loading relevant transactions 
library(dplyr)
library(lubridate)
# Identify the transactions table
transactions_table <- "core_transactions"

# Load the data from MySQL for a given day
date <- "2021-10-21"
transactions_df <- dbGetQuery(con, paste0("select * from ", transactions_table, " where trans_date = '", date, "'"))

# Filter the data with transaction amount > 1 million
transactions_df <- transactions_df %>% filter(amount > 1000000)

# Save the dataframe in a folder
saveRDS(transactions_df, file = paste0("transactions_", date, ".rds"))
```
 
#Rule2 R language
```{r}
con <-dbConnect(MySQL(),
                user="Kaimugul",
                password="Kabon@123#",
                dbname="Banking",
                host="127.0.0.1")
# identifying transactions table 
rule2<- "core_transactions"
rule2date<- "2021-10-21"
transactions_dfrule2 <- dbGetQuery(con, paste0("SELECT * FROM ", transactions_table, " where trans_date = '", date, "'"))

# Calculate average transaction value per customer
avg_transaction_per_customer <- transactions_dfrule2 %>%
  group_by(cust_id) %>%
  summarise(avg_transaction = mean(amount, na.rm = TRUE))


# Find total transaction value (debits) for each customer on the selected day
total_transaction_per_customer <- transactions_dfrule2 %>%
  filter(trans_date == rule2date) %>%
  group_by(cust_id) %>%
  summarise(total_transaction_value = sum(amount))

# Right join avg_transaction_per_customer and total_transaction_per_customer data.frames
merged_data <- right_join(avg_transaction_per_customer, total_transaction_per_customer, by = "cust_id")

# Calculate deviation percentage
merged_data <- merged_data %>%
  mutate(deviation_percentage = ((total_transaction_value - avg_transaction) * 100 / avg_transaction))

# Filter customers whose deviation is more than 100%
filtered_data <- merged_data %>%
  filter(deviation_percentage > 100)

saveRDS(transactions_dfrule2, file = paste0("transactionsrule2_", rule2date, ".rds"))
```
#Rule3 R language
```{r}
rule3date<-'2021-10-21'

# Load past 95 days data from the database
rule3 <- dbGetQuery(con, paste0("SELECT * FROM ", transactions_table, " WHERE trans_date >= '", rule3date, "'"))

#Convert the date columns to the proper date format
rule3$trans_date <- as.Date(rule3$trans_date)

# Check the first day of transaction & last day of transaction
transactions_rule3 <- rule3 %>%
  group_by(cust_id) %>%
  mutate(first_transaction_date = min(trans_date),
         last_transaction_date = max(trans_date)) %>%
  ungroup()

# Find number of days between first & last day
transactions_rule3 <- transactions_rule3 %>%
  mutate(days_inactive = as.numeric(last_transaction_date - first_transaction_date))

# If it is greater than 90 days, then save and report
inactive_customers_df <- transactions_rule3 %>%
  filter(days_inactive > 90)

# Save dataframe in a folder
saveRDS(inactive_customers_df, file = "inactive_customers.rds")

```
```{r}
# Load past 95 days data from the database
rule3 <- dbGetQuery(con, paste0("SELECT * FROM ", transactions_table, " WHERE trans_date >= '", rule3date, "'"))

# Convert the date columns to the proper date format
rule3$trans_date <- as.Date(rule3$trans_date)

# Check the first day of transaction & last day of transaction
transactions_rule3 <- rule3 %>%
  group_by(cust_id) %>%
  mutate(first_transaction_date = min(trans_date),
         last_transaction_date = max(trans_date)) %>%
  ungroup()

# Convert the first_transaction_date and last_transaction_date to proper date format
transactions_rule3$first_transaction_date <- as.Date(transactions_rule3$first_transaction_date)
transactions_rule3$last_transaction_date <- as.Date(transactions_rule3$last_transaction_date)

# Calculate the difference in months between the last transaction date and selected date
transactions_rule3$months_inactive <- as.numeric(difftime(rule3date, transactions_rule3$last_transaction_date, units = "days") / 30)

# Filter out customers who have been inactive for exactly 3 months
inactive_customers_df <- transactions_rule3 %>%
  filter(months_inactive == 3)

# Save dataframe in a folder
saveRDS(inactive_customers_df, file = "rule3.rds")

```

```{r}
# identifying transactions table 
rule4<- "core_transactions"
rule4date<- "2021-10-21"
transactions_df <- dbGetQuery(con, paste0("SELECT * FROM ", transactions_table, " where trans_date = '", date, "'"))
# Function to identify customers with high transaction volume for a given day
identify_high_transaction_volume_customers <- function(transactions_df, date) {
  # Filter transactions dataframe for the given day
  filtered_transactions <- transactions_df %>%
    filter(trans_date == as.Date(date))

  # Group transactions by customer ID and calculate total transaction amount for each customer
  total_transaction_amount_df <- filtered_transactions %>%
    group_by(cust_id) %>%
    summarize(total_transaction_amount = sum(amount))

  # Filter customers whose total transaction amount is greater than 1 million
  high_transaction_volume_customers <- total_transaction_amount_df %>%
    filter(total_transaction_amount > 1000000)

  return(high_transaction_volume_customers)
}

# Assuming you have loaded the transactions data into a dataframe named 'transactions_df'
# And you have provided the date for which you want to find high transaction volume customers
date <- "2023-07-24"

# Call the function to identify high transaction volume customers for the given date
high_transaction_volume_customers <- identify_high_transaction_volume_customers(transactions_df, date)

# Save the result in RDS format
saveRDS(high_transaction_volume_customers, file = paste0("high_transaction_volume_customers_", date, ".rds"))

# Print the list of high transaction volume customers for the given date
print(high_transaction_volume_customers)
```


#Rule4
```{r}
# identifying transactions table 
rule4<- "core_transactions"
rule4date<- "2021-10-21"
transactions_dfrule4 <- dbGetQuery(con, paste0("SELECT * FROM ", transactions_table, " where trans_date = '", date, "'"))

# Define the function to identify customers with high transaction volume
identify_customers_with_high_transaction_volume <- function(transactions_dfrule4, date) {
  # Filter the transactions dataframe to only include transactions on the given day
  filtered_transactions_df <- transactions_dfrule4 %>%
    filter(transaction_date == date)

  # Group the transactions by customer ID and calculate the total transaction amount for each customer
  total_transaction_amount_df <- filtered_transactions_df %>%
    group_by(cust_id) %>%
    summarize(total_transaction_amount = sum(amount))

  # Filter the results to only include customers whose total transaction amount is greater than 1 million
  high_transaction_volume_customers <- total_transaction_amount_df %>%
    filter(total_transaction_amount > 1000000)

  # Return the list of high transaction volume customers
  return(high_transaction_volume_customers)
}

# Identify customers with high transaction volume on the given day
high_transaction_volume_customers <- identify_customers_with_high_transaction_volume(transactions_dfrule4 = , "2023-10-21")

# Save the result in RDS format
saveRDS(high_transaction_volume_customers, file = "high_transaction_volume_customers.rds")

```

```{r}
con <-dbConnect(MySQL(),
                user="Kaimugul",
                password="Kabon@123#",
                dbname="Banking",
                host="127.0.0.1")
rule5<- "core_transactions"
rule5date<- "2021-10-21"
transactions_df <- dbGetQuery(con, paste0("SELECT * FROM ", rule5 table, " where trans_date = '", rule5date, "'"))

# Define the function to identify customers with high transaction volume over three consecutive days
identify_customers_with_high_transaction_volume <- function(transactions_df) {
  # Convert the transaction_date column to Date format
  transactions_df$trans_date <- as.Date(transactions_df$trans_date)

  # Sort the transactions data frame by cust_id and trans_date
  sorted_transactions_df <- transactions_df %>%
    arrange(cust_id, trans_date)

  # Group the transactions by cust_id and trans_date and calculate the cumulative transaction amount for each customer on each day
  cumulative_transactions_df <- sorted_transactions_df %>%
    group_by(cust_id, trans_date) %>%
    summarize(cumulative_amount = sum(amount)) %>%
    ungroup()

  # Create a window of three consecutive days for each customer
  windowed_transactions_df <- cumulative_transactions_df %>%
    group_by(cust_id) %>%
    mutate(window_sum = cumsum(cumulative_amount) - lag(cumulative_amount, default = 0))

  # Filter the results to include customers whose cumulative transaction amount is greater than 1 million over three consecutive days
  high_transaction_volume_customers <- windowed_transactions_df %>%
    filter(window_sum > 1000000)

  # Return the list of high transaction volume customers
  return(high_transaction_volume_customers)
}

# Assuming you have a data frame named transactions_df containing columns: cust_id, trans_date, amount

# Identify customers with high transaction volume over three consecutive days
high_transaction_volume_customers <- identify_customers_with_high_transaction_volume(transactions_df)

# Save the result in RDS format
saveRDS(high_transaction_volume_customers, file = "rule51.rds")

```







```


#Rule2
```{R}
con <-dbConnect(MySQL(),
                user="Kaimugul",
                password="Kabon@123#",
                dbname="Banking",
                host="127.0.0.1")
#Query the MYSQL table
query<-"WITH avg_transaction_value AS (
   SELECT cust_id, AVG(amount) AS avg_transaction_value
   FROM core_transactions
   GROUP BY cust_id
)
SELECT core_transactions.cust_id, amount,
       ABS(amount - avg_transaction_value) * 100 AS deviation_percentage
FROM core_transactions
JOIN avg_transaction_value ON core_transactions.cust_id = avg_transaction_value.cust_id
WHERE ABS(amount - avg_transaction_value) * 100 > 100;"
results21<-dbGetQuery(con, query)
print(results21)

dbDisconnect(con)
```
#Rule3
#rule4
```{r}
con <-dbConnect(MySQL(),
                user="Kaimugul",
                password="Kabon@123#",
                dbname="Banking",
                host="127.0.0.1")
# Query the MySQL table
query <- "select total_daily_transactions, cust_id
from 
        (SELECT cust_id, trans_date, count(*) AS num_transactions, sum(amount) as total_daily_transactions
          FROM core_transactions
          GROUP BY cust_id, trans_date) AS daily_transaction_summary
          where total_daily_transactions > 1000000
          and trans_date = '2021-10-31'
          limit 10;"

result4 <- dbGetQuery(con, query)
print(result4)

dbDisconnect(con)
```
#Rule5
```{r}
con <-dbConnect(MySQL(),
                user="Kaimugul",
                password="Kabon@123#",
                dbname="Banking",
                host="127.0.0.1")
# Query the MySQL table
query <- "SELECT cust_id, total_transactions
FROM (
    SELECT cust_id, sum(amount) AS total_transactions, date(trans_date) AS transaction_date
    FROM core_transactions
    GROUP BY cust_id, transaction_date
    ORDER BY cust_id, transaction_date
) AS transaction_summary
WHERE total_transactions > 1000000 
  AND transaction_date BETWEEN DATE_SUB('2023-04-04', INTERVAL 3 DAY) AND '2023-04-04';
"
result5 <- dbGetQuery(con, query)
print(result5)

dbDisconnect(con)
```
#Rule6
```{r}
con <-dbConnect(MySQL(),
                user="Kaimugul",
                password="Kabon@123#",
                dbname="Banking",
                host="127.0.0.1")
# Query the MySQL table
query <- "SELECT cust_id, customer_transactions
FROM (
    SELECT cust_id, sum(amount) AS customer_transactions, date(trans_date) AS transaction_date
    FROM core_transactions
    GROUP BY cust_id, transaction_date
    ORDER BY cust_id, transaction_date
) AS transaction_summary
WHERE customer_transactions > 2000000 
  AND transaction_date BETWEEN DATE_SUB('2022-02-07', INTERVAL 1 week) AND '2022-02-07';
"
result6 <- dbGetQuery(con, query)
print(result6)

dbDisconnect(con)
```





#Rule7
```{r}
con <-dbConnect(MySQL(),
                user="Kaimugul",
                password="Kabon@123#",
                dbname="Banking",
                host="127.0.0.1")
# Query the MySQL table
query <- "WITH average_monthly_transactions AS (
    SELECT
        cust_id,
        AVG(amount) AS monthly_transaction
    FROM core_transactions
    WHERE trans_date >= '2023-06-25' - INTERVAL 6 MONTH
    GROUP BY cust_id
),
current_month_transactions AS (
    SELECT
        cust_id,
        SUM(amount) AS total_transaction_amount
    FROM core_transactions
    WHERE trans_date >= '2023-06-25'
    GROUP BY cust_id
)
SELECT
    c1.cust_id,
    total_transaction_amount,
    monthly_transaction,
    ((total_transaction_amount - monthly_transaction) * 100 / monthly_transaction) AS deviation_percentage
FROM current_month_transactions AS c1
JOIN average_monthly_transactions AS c2 ON c1.cust_id = c2.cust_id
WHERE ((total_transaction_amount - monthly_transaction) * 100 / monthly_transaction) > 100;"
results1 <- dbGetQuery(con, query)
print(results1)

dbDisconnect(con)
```

```{R}
con <-dbConnect(MySQL(),
                user="Kaimugul",
                password="Kabon@123#",
                dbname="Banking",
                host="127.0.0.1")
#loading R libraries
library(dplyr)
library(lubridate)
#selecting my month
month<-'2023-06'
#identifying total transactions customers did 
```